### 2020-02-22 main_text/main_text.tex
Another round of edits by KL

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.4\textwidth]{media/Fig1_data.pdf}
		\caption{{\bf Experimental data on the \emph{xap} circuit.} The expression of the \emph{xapAB} promoter was measured for different concentrations of xanthosine in the bacteria's environment. (A) shows the wild type circuit while in (B) the genes \emph{xapABR} and in (C) the genes \emph{xapAB} are removed. The wild type circuit behaves like a switch.}
		\label{fig1:data}
	\end{figure}

@@

	Clearly, XapR regulates the \emph{xapAB} promoter. As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site, and one further upstream. When the upstream binding site is removed, bimodality is lost and the expression of \emph{xapAB} becomes several orders of magnitude weaker, as can be seen in Fig~\ref{}.
	\textcolor{red}{data missing}

	Already, we believe this data, combined with the prior knowledge of the field, clearly suggest a minimal model of the \emph{xapABR} system. The construction and exploration of this model occupies the remainder of this work. We leave it for future work to quantitatively dissect and test the model, in the manner of, for example, \textcolor{red}{citations}.

@@

	\begin{figure}[h!]
		\centering
		\includegraphics{media/Fig2_model.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB is treated as a Michaelis-Menten enzyme which imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	\subsection*{Deterministic phase portraits}
	A standard way to analyze dynamical systems deterministically is
	to plot phase portraits. In the following, we present such plots where the
	state variables are the mRNA, the protein, and the xanthosine concentration. Note that for the low copy numbers that can occur in our system, a deterministic analysis is not necessarily meaningful. Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results. Thus, deterministic phase portraits are a good starting point.

@@

	\paragraph*{The extracellular xanthosine concentration.}
	The parameter that is the experimentally most easily tunable and
	biologically the most relevant is the extracellular xanthosine
	concentration. When it is increased in experiments, the cells go from (1.) all
	being in the low expression state to (2.) the population being in a mixed state with some cells in a low expression state and others in a high expression state (all-or-none phenomenon) and then to (3.) all being in the high
	expression state (see section ``experimental motivation''). If our model is correct, it should
	exhibit the same qualitative behavior. Indeed we find exactly this: as can
	be seen in Fig~\ref{fig5:extraxanth}, increasing $\n{[c]_a}$ makes the high
	stable fixed point appear and then, for even higher $\n{[c]_a}$, the lower
	one disappears. Thus, for low $\n{[c]_a}$ the only stable point of the system
	is at low expression, and for high $\n{[c]_a}$ there is only high
	expression. In between, there are two deterministically stable expression
	levels. Later on in this text, we show this result in form of a bifurcation diagram.

@@

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. Also note the similarity to the shape of the experimental distributions that were presented in section ``experimental motivation''. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This actually is a good finding: the biologically interesting feature is the switch-like behavior, whereas bimodality can be problematic for biology as this means that there is a large insecurity in the input-output behavior of a bacteria's reaction to xanthosine. Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch.

	We found that the two lower fixed points (marked as 1 and 2 in
	Fig~\ref{fig4:bistability}) need to be very close like in
	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
	meaning larger distance between the first and second fixed point,
	almost no switching was observed. Of course, switching is also a matter of the
	waiting time and stochastic effects: if one waits for long enough, it
	should eventually occur. However, switching times of more than several hours
	are not at the center of this investigation and would mean that switching is
	extremely unlikely. There are two aspects that become relevant in this
	context that we neglect in our analysis but briefly mention here:
	transcription and translation bursts lead to higher stochasticity and cell
	division leads to some discontinuity in the process.

	Note that while the deterministic analysis assumes the variables to be
	continuous, the simulations work with discrete numbers of mRNA, protein, and
	xanthosine. This per se is no problem, because the deterministic analysis
	describes the mean values and the simulation fluctuates around this mean.
	If, however, the mRNA number at the third (high) fixed point is too low, the
	system will not be able to resolve the two points anymore. It is too low when the distance between the first and the third fixed point
	becomes as low as the stochastic fluctuations in the system, which is around
	3 mRNA molecules. In Fig~\ref{fig8:stochC}, the mRNA number at the low fixed
	point was fluctuating between 0 and 2 with a mean significantly smaller than
	1, and at the high fixed point it was varying between 2 and 6.

	\paragraph*{Time evolution and adaptation times.}
	Fig~\ref{fig9:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
	Again, the simulation was started with a mRNA, protein and intracellular
	xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
	\unit{s}$. In this specific example, the fixed point was reached after $10^5
	\unit{s}$. Of course, the system moves to its mean steady-state much faster when the extracellular xanthosine concentration is further away from the bimodal regime (in analogy to a ``critical point''). For comparison, the trajectory obtained from integrating the deterministic system is also shown in Fig~\ref{fig9:StochT}. It clearly agrees well with the simulation. There are some cases where the big adaptation step happens slightly earlier or later, but still at around $10^5 \unit{s}$.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB).}
			Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
			at an mRNA, protein and intracellular xanthosine count of 0. In this
			case the system reached the high-expression state after $10^5
			\unit{s}$. The parameters that were used are the same as in
			Table~\ref{table1:nondim}, the only exception being the
			extracellular xanthosine concentration, which was chosen to be
			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
			Fig~\subref{fig8:stochC}{(C)}.}
		\label{fig9:StochT}
	\end{figure}

@@

	Nevertheless, we do warn the reader that the timescales in the simulations and even more so in the deterministic system
	should be taken with reservation. Fluctuations in the parameters are not considered here, and neither are cell divisions or the burstiness of transcription and translation. This means
	that stochasticity may be larger in the real system which should have an
	influence on the timescales and may shorten the time until the fixed point is reached. However, including transcriptional bursts in the stochastic simulation changes little in the output: the qualitative behavior remains the same but fluctuations around the mean as well as in the adaptation time become larger and bimodality already occurs at lower $\mathrm{[c]_a}$ (see~\nameref{S1_Text}). Still, the fact that there are no big changes shows that the system is stable to stochastic perturbations and our assumptions should not be too significant.

	Note that in a biologically realistic setting, the extracellular xanthosine concentration would probably be high enough to be in the regime where there only is the upper fixed point and thus no bistability or even bimodality. As mentioned before, the system reaches its mean steady-state faster in that regime, and hence, the bacteria can adapt more quickly.

@@

	The distributions we obtained can be found in~\nameref{S1_Text}. Here, we instead present the results in form of a bifurcation diagram in Fig~\ref{fig10:Bifurcation}. In blue, the positions of the deterministic fixed points for the corresponding extracellular xanthosine concentration are shown. As explained before, there is only one fixed point for low and high values of $\mathrm{[c]_a}$, but in between, there are three. In yellow and orange, the results from the stochastic simulation are shown, where the mean of the resulting distribution was taken. For the orange points, the simulation was started at $\mathrm{[m]_a}=\mathrm{[p]_a}=\mathrm{[x]_a}=0$, leading to the distributions in Fig~\ref{fig8:stochC}. In contrast, the yellow points show the results from starting the simulation at the $\mathrm{[m]_a}$, $\mathrm{[p]_a}$, and $\mathrm{[x]_a}$ values of the high fixed point. Note that by taking the mean of each distribution, we hide the feature of bimodality. For that reason, we additionally show the position of the two peaks in the bimodal distribution by splitting it into two parts at its minimum and calculating the mean of each of the parts. Also note that it suffices to show the bifurcation diagram in one dimension (here: protein concentration) because the fixed points are uniquely determined by that (recall Fig~\ref{fig4:bistability}, the fixed points are unique points in the 3D system).

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig10_bifurcation.eps}
		\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
			The blue line shows the positions of deterministic fixed points, where the dashed line indicates the instability of the middle one. The orange and yellow points show the mean in the stochastic simulations when started at zero and at the high fixed point, respectively. The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines. The arrows illustrate the hysteresis loop in the stochastic simulations.}
		\label{fig10:Bifurcation}
	\end{figure}

@@

	In addition to the hysteresis, Fig~\ref{fig10:Bifurcation} also illustrates the astonishingly good agreement of the stochastic simulation and the deterministic results, despite having copy numbers below 10 in some cases. Although one should be cautious about this because of higher stochasticity in the real biological system (addressed in the previous paragraph), the result does show that the switch-like feature in the circuit is strong and stable.

@@

	Phase diagrams, showing for which parameters the system is bistable and for
	which there is only the lower or the upper stable fixed point, could be
	calculated from arguments made in~\cite{Cherry2000}. However, the
	simulations showed that in the deterministically bistable regime, it is dependent on initial conditions which fixed points the system occupies, which is why we have refrained from showing such phase
	diagrams. Furthermore, the timescales in the problem could be investigated
	more thoroughly, for example the dependence of the switching time on
	$\n{[c]_a}$, but such an analysis should probably account for	cell divisions and fluctuations in the parameters, which is not straightforward. Lastly, it could be interesting to investigate the magnitude of the fluctuations around the fixed point away from and near the bifurcations in $\mathrm{[c]_a}$.

@@

@@

		\bibitem{RazoMejia2018}
		Razo-Mejia M, Barnes SL, Belliveau NM, Chure G, Einav T, Lewis M, et~al.
		\newblock Tuning Transcriptional Regulation through Signaling: A Predictive
		Theory of Allosteric Induction.
		\newblock Cell Systems. 2018;6(4):456 -- 469.e10.
		\newblock doi:{10.1016/j.cels.2018.02.004}.

>>>>

	\begin{figure}%[h!]
		\centering
		\includegraphics[width=0.4\textwidth]{media/Fig1_data.pdf}
		\caption{{\bf Experimental data on the \emph{xap} circuit.} The expression of the \emph{xapAB} promoter was measured for different concentrations of xanthosine in the bacteria's environment. (A) shows the wild type circuit while in (B) the genes \emph{xapABR} and in (C) the genes \emph{xapAB} are removed. The wild type circuit behaves like a switch. \textcolor{red}{placeholder figure up to now}}
		\label{fig1:data}
	\end{figure}

@@

	Clearly, XapR regulates the \emph{xapAB} promoter. As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site and one further upstream. When the upstream binding site is removed, bimodality is lost and the expression of \emph{xapAB} becomes several orders of magnitude weaker, as can be seen in Fig~\ref{}.
	\textcolor{red}{data missing}

	Already, we believe this data, combined with the prior knowledge of the field, clearly suggest a minimal model of the \emph{xapABR} system. The construction and exploration of this model occupies the remainder of this work. We leave it for future work to quantitatively dissect and test the model, in the manner of~\cite{Garcia2011,RazoMejia2018,Chure2019}, for example.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig2_model.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB is treated as a Michaelis-Menten enzyme which imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	\subsection*{Deterministic phase portraits}
	A standard way to analyze dynamical systems deterministically is
	to plot phase portraits. In the following, we present such plots where the
	state variables are the mRNA, the protein, and the xanthosine concentration. Note that for the low copy numbers that can occur in our system, a deterministic analysis is not necessarily meaningful. Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results. Thus, deterministic phase portraits are a surprisingly good starting point.

@@

	\paragraph*{The extracellular xanthosine concentration.}
	The parameter that is the experimentally most easily tunable and
	biologically the most relevant is the extracellular xanthosine
	concentration. When it is increased in experiments, the cells go from (1.) all
	being in the low expression state to (2.) the population being in a mixed state with some cells in a low expression state and others in a high expression state (all-or-none phenomenon) and then to (3.) all being in the high
	expression state (see section ``experimental motivation''). If our model is correct, it should
	exhibit the same qualitative behavior. Indeed we find exactly this: as can
	be seen in Fig~\ref{fig5:extraxanth}, increasing $\n{[c]_a}$ makes the high
	stable fixed point appear and then, for even higher $\n{[c]_a}$, the lower
	one disappears. Thus, for low $\n{[c]_a}$ the only stable point of the system
	is at low expression, and for high $\n{[c]_a}$ there is only high
	expression. In between, there are two deterministically stable expression
	levels. Another means of visualizing this is with an bifurcation diagram, which we discuss in detail later.

@@

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This actually is a good finding: it implies that bistability, characterized by two separate \emph{stable} steady-states, largely persists even in the presence of stochastic fluctuations. Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch and the deterministic picture is sufficient except in a small range of parameter space. Whether or not that is where the system actually is remains a question for future experiments.

	We found that the two lower fixed points (marked as 1 and 2 in
	Fig~\ref{fig4:bistability}) need to be very close like in
	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
	meaning larger distance between the first and second fixed point,
	almost no switching was observed. Of course, switching is also a matter of the
	waiting time and stochastic effects: if one waits for long enough, it
	should eventually occur. However, switching times of more than several hours
	are not at the center of this investigation and would mean that switching is
	extremely unlikely. There are two aspects that become relevant in this
	context that we neglect in our analysis but briefly mention here:
	transcription and translation bursts lead to higher stochasticity and cell
	division leads to some discontinuity in the process. The effect of bursts is addressed in~\nameref{S1_Text}.

	Note that while the deterministic analysis assumes the variables to be
	continuous, the simulations work with discrete numbers of mRNA, protein, and
	xanthosine. This per se is no problem, because the deterministic analysis
	describes the mean values and the simulation fluctuates around this mean.
	If, however, the mRNA number at the third (high) fixed point is too low, the
	system may not be able to resolve the two points anymore. The tolerance to this is surprisingly large, though: In Fig~\ref{fig8:stochC}, the distance between the first and the third fixed point is around 3 mRNA molecules. While this is around the size of the fluctuations in Fig~\ref{fig8:stochC}, the latter become as large as 10 mRNA molecules when bursts are included in the simulation (see~\nameref{S1_Text}) and yet, the system is able to resolve the fixed points very well.

	\paragraph*{Time evolution and adaptation times.}
	Fig~\subref{fig9:StochT}{A} shows the time evolution of the protein concentration for one typical run of the simulation.
	Again, the simulation was started with a mRNA, protein and intracellular
	xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
	\unit{s}$. In this specific example, the fixed point was reached after roughly $1.5 \cdot 10^5
	\unit{s}$ (adaptation time). This time varies: Fig~\subref{fig9:StochT}{B} shows the distribution of adaptation times from 1000 runs of the simulation. In both figures, the trajectory obtained from integrating the deterministic system is shown for comparison. It clearly agree well with the simulation on average.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
			Table~\ref{table1:nondim}, the only exception being the
			extracellular xanthosine concentration, which was chosen to be
			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
			Fig~\subref{fig8:stochC}{(C)}. (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)). The orange line shows the corresponding deterministic time. The green dashed line right next to it indicated the mean of the blue distribution.}
		\label{fig9:StochT}
	\end{figure}

@@

	Nevertheless, we do warn the reader that the timescales in the simulations and even more so in the deterministic system
	should be taken with reservation. Fluctuations in the parameters are not considered here, and neither are cell divisions or the burstiness of transcription and translation. This means
	that stochasticity may be larger in the real system which should have an
	influence on the timescales and may shorten the time until the fixed point is reached. However, including transcriptional bursts in the stochastic simulation changes little in the output: the qualitative behavior remains the same but fluctuations around the mean as well as in the adaptation time become larger and bimodality already occurs at lower $\mathrm{[c]_a}$ (see~\nameref{S1_Text}). Still, the fact that there are no big changes shows that the system is stable to stochastic perturbations and our particular assumptions should not be too significant.

	Of course, the system moves to its mean steady-state much faster when the extracellular xanthosine concentration is further away from the bimodal regime (in analogy to a ``critical point''). 
	Note that it could well be that in reality, the extracellular xanthosine concentration is high enough to be in the regime where there only is the upper fixed point and thus no bistability or even bimodality. As the system reaches its mean steady-state faster in that regime, the bacteria could adapt more quickly.

@@

	The distributions we obtained can be found in~\nameref{S1_Text}. Here, we instead present the results in the form of a bifurcation diagram in Fig~\ref{fig10:Bifurcation}. In blue, the positions of the deterministic fixed points for the corresponding extracellular xanthosine concentration are shown. As explained before, there is only one fixed point for low and high values of $\mathrm{[c]_a}$, but in between, there are three. In yellow and orange, the results from the stochastic simulation are shown, where the mean of each distribution was taken. For the orange points, the simulation was started at $\mathrm{[m]_a}=\mathrm{[p]_a}=\mathrm{[x]_a}=0$, leading to the distributions in Fig~\ref{fig8:stochC}. In contrast, the yellow points result from starting the simulation at the $\mathrm{[m]_a}$, $\mathrm{[p]_a}$, and $\mathrm{[x]_a}$ values of the high fixed point. Note that by taking the mean of each distribution, we hide the feature of bimodality. For that reason, we additionally show the approximate position of the two peaks in the bimodal distribution. We do this by estimating the position of the minimum between the two peaks of the distribution, splitting it into two parts there, and calculating the mean of each of the parts. Also note that it suffices to show the bifurcation diagram in one dimension (here protein concentration) because this uniquely determines the fixed points (recall from Fig~\ref{fig4:bistability}, the fixed points are unique points in the 3D system).

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig10_bifurcation.eps}
		\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
			The blue line shows the positions of deterministic fixed points, while the dashed line indicates the instability of the middle one. The orange and yellow points show the mean in the stochastic simulations when started at zero and at the high fixed point, respectively. The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines. The arrows illustrate the hysteresis loop in the stochastic simulations.}
		\label{fig10:Bifurcation}
	\end{figure}

@@

	In addition to the hysteresis, Fig~\ref{fig10:Bifurcation} also illustrates the astonishingly good agreement of the stochastic simulation and the deterministic results, despite having copy numbers below 10 in some cases. Although one should be cautious about this because of higher stochasticity in the real biological system (addressed above), the result does show that the switch-like feature in the circuit is strong and stable.

@@

	Phase diagrams, showing for which parameters the system is bistable and for
	which there is only the lower or the upper stable fixed point, could be
	calculated from arguments made in~\cite{Cherry2000}. However, the
	simulations showed that in the deterministically bistable regime, which fixed point(s) the system occupies is dependent on initial conditions, which is why we have refrained from showing such phase
	diagrams. Furthermore, the timescales in the problem could be investigated
	more thoroughly, for example the dependence of the switching time on
	$\n{[c]_a}$, but such an analysis should probably account for	cell divisions and fluctuations in the parameters, which is not straightforward. Lastly, it could be interesting to investigate the magnitude of the fluctuations around the fixed point away from and near the bifurcations in $\mathrm{[c]_a}$.

@@

		\bibitem{Garcia2011}
		Garcia HG, Phillips R.
		\newblock Quantitative dissection of the simple repression input-output
		function.
		\newblock Proceedings of the National Academy of Sciences of the United States
		of America. 2011;108 29:12173--8.

		\bibitem{RazoMejia2018}
		Razo-Mejia M, Barnes SL, Belliveau NM, Chure G, Einav T, Lewis M, et~al.
		\newblock Tuning Transcriptional Regulation through Signaling: A Predictive
		Theory of Allosteric Induction.
		\newblock Cell Systems. 2018;6(4):456 -- 469.e10.
		\newblock doi:{https://doi.org/10.1016/j.cels.2018.02.004}.

		\bibitem{Chure2019}
		Chure G, Kaczmarek ZA, Phillips R.
		\newblock Physiological Adaptability and Parametric Versatility in a Simple
		Genetic Circuit.
		\newblock bioRxiv. 2019;doi:{10.1101/2019.12.19.878462}.

@@

### 2020-02-22 s1_text/s1_text.tex
Another round of edits by KL

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{FigSI5.pdf}
	\caption{{\bf A family of curves for increasing $\n{[c]_a}$.}
		The blue curve is the protein nullcline, which does not change with $\n{[c]_a}$. The other lines are the xanthosine nullclines for different values of the extracellular xanthosine concentration.}
	\label{figS5:incC}
\end{figure}

\paragraph*{Including transcriptional bursting.} To test the robustness of our results to larger stochasticity (which is expected in the real biological system), we included some burstiness in transcription in our stochastic simulations. We assumed a geometric distribution for the burst size $s$, i.e. $p(s) = (1/4)\cdot (1-1/4)^{s-1}$ with mean $4$, which is a bit larger than what usually occurs in biology \textcolor{red}{citation missing}. We then cut this distribution at a burst size of 10, meaning we did not consider larger bursts, leading to an actual mean of around $3.2$. The transcription rate was rescaled such that the mean number of mRNA's produced remained the same.

We observe the same qualitative behavior but with stronger fluctuations around the mean and also larger variation in the time to the steady-state mean. As an example, Fig~\ref{evolbursts} shows the time evolution of two runs of the simulation with bursting for the same conditions (i.e. parameters and starting values) as the time evolution that was shown in the main text. Fig~\ref{distbursts} shows the distribution obtained for the same and an additional condition (i.e. parameters and starting values) as for the distributions shown in the main text. The variance in the timescales and distributions is larger than without bursting. Furthermore, switching already happens for lower $\mathrm{[c]_a}$: in \subref{distbursts}{C}, most cells are in the high expression state, whereas in the corresponding distribution in the main text, there were many uninduced cells, i.e. the first peak was higher. To obtain a bimodal distribution like in the main text, $\mathrm{[c]_a}$ now needs to be set to $17.5$ instead of $18.5$ as shown in \subref{distbursts}{B}. This was to be expected, since the larger fluctuations make it easier for the system to jump across the middle fixed point.

Still, there are no big qualitative differences to the results without bursting. Thus, the assumptions we made that lead to less stochasticity than in the real biological system are not expected to have a very significant influence on our results, and the switch-like behavior seems to be stable to stochastic perturbations. \textcolor{red}{Fig 7 C not enough runs yet}

\begin{figure}
	\centering
	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
	\caption{{\bf Time evolution from the stochastic simulation when transcriptional bursting is included.}
		(A) and (B) each show the time evolution from one run of the simulation. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
		at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in the
		table in the main text, the only exception being the
		extracellular xanthosine concentration, which was chosen to be
		$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).}
	\label{evolbursts}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1\textwidth]{FigSI6_7_2.pdf}
	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
		(A) shows the distribution obtained from the simulation with $\n{[c]_a}=12$, (B), (C), and (D) show the same for $\n{[c]_a}=17.5$, $\mathrm{[c]_a}=18.5$, and $\mathrm{[c]_a}=25$, respectively (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
		was run 1000 times for $10^6 \unit{s}$ each (simulated time) and
		started at a mRNA, protein and intracellular xanthosine count of 0.}
	\label{distbursts}
\end{figure}

@@

@@

>>>>

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{FigSI5.pdf}
	\caption{{\bf A family of curves for increasing $\n{[c]_a}$.}
		The blue, dashed curve is the protein nullcline, which does not change with $\n{[c]_a}$. The other curves are the xanthosine nullclines for different values of the extracellular xanthosine concentration.}
	\label{figS5:incC}
\end{figure}

\paragraph*{Including transcriptional bursting.} To test the robustness of our results to larger stochasticity (which is expected in the real biological system), we explored how the addition of burstiness in transcription altered our stochastic simulations. We assumed a geometric distribution for the burst size $s$, i.e. $p(s|\theta) = \theta\cdot (1-\theta)^{s-1}$ with mean $\theta^{-1}=4$, which is comparable to or slightly larger than reported values~\cite{Golding2005}. We then truncate this distribution at a burst size of 10, meaning we did not consider larger bursts, leading to an actual mean of around $3.2$. The transcription rate was rescaled such that the mean number of mRNA's produced remained the same.

We observe the same qualitative behavior but with stronger fluctuations around the mean and larger variation in the time to reach the steady-state mean. As an example, Fig~\subref{evolbursts}{A} shows the time evolution of one run of the simulation with bursting for the same conditions (i.e. parameters and starting values) as the time evolution that was shown in the main text. In addition, the distribution of adaptation times can be seen in Fig~\subref{evolbursts}{B}. Fig~\ref{distbursts} shows the distribution obtained for the same and an additional condition (i.e. parameters and starting values) as for the distributions shown in the main text. The variance in the timescales and distributions is larger than without bursting. Furthermore, switching already happens for lower $\mathrm{[c]_a}$: in \subref{distbursts}{C}, most cells are in the high expression state, whereas in the corresponding distribution in the main text, there were many uninduced cells, i.e. the first peak was higher. To obtain a bimodal distribution like in the main text, $\mathrm{[c]_a}$ now needs to be set to $17.5$ instead of $18.5$ as shown in \subref{distbursts}{B}. This was to be expected, since the larger fluctuations make it easier for the system to jump across the middle fixed point.

Still, there are no big qualitative differences to the results without bursting. Thus, the assumptions we made that lead to less stochasticity than in the real biological system are not expected to have a very significant influence on our results, and the switch-like behavior seems to be stable to stochastic perturbations.

\begin{figure}
	\centering
	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
	\caption{{\bf Time evolution and adaptation times from the stochastic simulation when transcriptional bursting is included.}
		(A) shows the time evolution from one run of the simulation. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
		at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in the
		table in the main text, the only exception being the
		extracellular xanthosine concentration, which was chosen to be
		$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 200 runs of the simulation (same conditions as (A)). The distribution from the main text (without bursting) is shown in orange for comparison. \textcolor{red}{will add some more runs, but the simulation takes forever}}
	\label{evolbursts}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1\textwidth]{FigSI6_7_2.pdf}
	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
		Blue shows the distributions from the simulation with bursts, orange shows those without bursts from the main text for comparison. The following values of $\mathrm{[c]_a}$ were used: (A) $\n{[c]_a}=12$, (B) $\n{[c]_a}=17.5$ (not shown in main text), (C) $\mathrm{[c]_a}=18.5$, and (D) $\mathrm{[c]_a}=25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
		was run 1000 times for $10^6 \unit{s}$ each (simulated time) and
		started at a mRNA, protein and intracellular xanthosine count of 0.}
	\label{distbursts}
\end{figure}

@@

Note that in the case with bursts that is discussed above, the transcription term $r_{\n{m}} p_{\n{active}}(x) P(m-1,p,x)$ is changed to $r_{\n{m}} p_{\n{active}}(x) \sum_{k=1}^{10} P(m-k,p,x) \frac{\theta\cdot (1-\theta)^{k-1}}{Z}$ with $Z = \sum_{k=1}^{10} k \cdot \theta\cdot (1-\theta)^{k-1}$ and $\theta = 1/4$ in our example.

@@

	\bibitem{Golding2005}
	Golding I, Paulsson J, Zawilski SM, Cox EC.
	\newblock Real-Time Kinetics of Gene Activity in Individual Bacteria.
	\newblock Cell. 2005;123(6):1025--1036.
	\newblock doi:{10.1016/j.cell.2005.09.031}.
